A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-01 latex/main.tex

[TODO narrative is muddled here: the point of the para is to say that 
pango X lineage designation is manual and subjective.]
% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
% ‘Pangolin’ nomenclature system. "
Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
can be attributed to the clarity of the rules under which new lineages 
are designated and their explicitly phylogenetic basis. 
The first recombinant lineage XA was designated in 2021 based on the 
analysis of Jackson et al.\cite{Jackson2021}, which established 
that recombination with onward transmission had occured [weak FIXME].
As of pango-designation version v1.33.1 (2025-03-31), a total of 146 distinct 
X lineages have been designated.
%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
% 146
Although the process of incorporating recombinant lineages in the Pango 
nomenclature system is well defined,
the identification of such events is much more subjective, requiring
the synthesis of multiple lines of evidence and often the 
manual inspection of phylogenies and sequence alignments
by volunteers.

@@

\subsection*{Quality control of recombination events}
Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions \cite{DeMaio2020issues} (MNSs)
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
We developed a quality control (QC) measure 
which counts the number of ``loci'' (clusters of closely spaced sites)
supporting the left and right parents of a recombinant (STAR Methods),
and effectively identifies a range of issues. 
Figure~\ref{fig:recombinant_qc} plots the net number of supporting
loci for the left and right parents for all 855 recombination
events. The 501 recombinants with fewer than 4 net loci supporting 
each parent are highly enriched for the AmpliSeq V1 sequencing
protocol [TODO concrete numbers?], 
and 71 (14\%) are associated with 6-base deletion in the 
Delta lineage, known to be affected 
by amplicon dropout\cite{Sanderson2021}.
The 354 remaining recombination events that pass this QC filter
contain all of the recombinants associated with Pango X lineages
(Table~\ref{tab:pango_x_lineages}) and 
the Jackson et al recombinants (Document\ref{sec:jackson_recombs}).

To assist in the identification of such issues
we developed a SnpIt-like~\cite{OToole2024} visualisation 
that shows the informative genomic positions between 
a recombinant and its parent sequences, and highlights 
potentially problematic loci.
Each copying pattern shows the positions
where the allelic state in a recombinant
(middle row) matches that of the left (upper row, coloured green) 
or right parent (lower row, coloured blue),
or where the recombination event requires a de-novo mutation
(gold, with mutational change below).
See Figures~\ref{fig:averted_mutations}D and
\ref{fig:pango_x_copying_patterns} for copying patterns of the 
Pango X lineages, and 
Figure~\ref{fig:copying_pattern_examples} for copying patterns
highlighting specific types of QC problem (and more details on the 
visualisation).

@@

The LS HMM at the heart of sc2ts provides a simple and powerful means 
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution for the 
putative recombinant sequence with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between 
the two solutions---the number of 
mutations ``averted'' by recombination---allows us to quantify
the plausibility.

Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination,
as measured by the number of resulting mutations 
The more mutations are ``averted'' by allowing recombination,
the stronger the evidence for recombination.
We calculated the number of averted mutations for
all the ``robust'' recombination events (STAR Methods),
which include all the Pango X-associated recombination events
(Table~\ref{tab:pango_x_lineages})
and those studied by Jackson et al. \cite{Jackson2021}
(Table \ref{tab:jackson}, Document~\ref{sec:jackson_recombs}).

Our analysis of the number of averted mutations reveals a spectrum of plausibility
across all the ``robust'' recombination events
(Figure~\ref{fig:averted_mutations}A).
Examples of highly plausible recombinants are XA and XBB
(for an extended analysis, see Document~\ref{sec:supplement_indepth_pango_x_lineages}).
The ``copying pattern'' of XA visually confirms
that a high number of recombination-informative sites support the suggested parents 
(Figure~\ref{fig:averted_mutations}D).
XW is an example of a marginally well-supported recombinant.
The initial evidence for proposing XW for designation is four nucleotides associated with the suggested BA.1 parent.
% FIXME What's document S3?
The copying patterns of all the recombination events are shown in Document S3.

[FIXME - discuss BA.5 (kept as a recombination event but only because we can't 
rewire it post-hoc, owning to subsequent tree building) and the BA.2 event (doesn't
pass QC.]
[Our results suggest that Delta and Omicron BA.2 might have recombinant origins [no longer true],
but we interpret these results with caution,
as these VOCs are undersampled around the time of their emergence (REF).]
Note that the presence of many de-novo mutations
observed in the copying patterns of XBB (Figure 3D) suggests the existence of
many genetically distinct, unsampled ancestors [FIXME: in this case, probably BJ.1, see
https://github.com/jeromekelleher/sc2ts-paper/issues/336].
Future work is needed to test whether these VOCs descended from recombination events,
with improved sampling, if possible.

@@

% https://github.com/jeromekelleher/sc2ts-paper/issues/717
167 (43\%) of QC-passing recombination events were also detected as recombinants by rebar.
We hypothesize that this disagreement arose from
differences in sensitivity between these methods.
Consistent with this, we found that rebar needs a higher level of divergence between parents
to correctly classify a sequence as recombinant, as shown by
a shorter time to their MRCA (Figure \ref{fig:averted_mutations}B) and
a lower distance between their Pango identity (Figure \ref{fig:averted_mutations}C).

@@

We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene (Figure~\ref{fig:recombinant_intervals}A,B),
consistent with patterns observed in previous studies
\cite{Bobay2020recombination,Turakhia2022,Li2024CovRecomb}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
allowing recombination endpoints to be estimated more precisely.

>>>>

[TODO narrative is muddled here: the point of the para is to say that 
pango X lineage designation is manual and subjective.]
% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
% ‘Pangolin’ nomenclature system. "
Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
can be attributed to the clarity of the rules under which new lineages 
are designated and their explicitly phylogenetic basis. 
The first recombinant lineage XA was designated in 2021 based on the 
analysis of Jackson et al.\cite{Jackson2021}, which established 
that recombination with onward transmission had occured [weak FIXME].
As of pango-designation version v1.33.1 (2025-03-31), a total of 146 distinct 
X lineages have been designated.
%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
% 146
Although the process of incorporating recombinant lineages in the Pango 
nomenclature system is well defined,
the identification of such events is much more subjective, requiring
the synthesis of multiple lines of evidence and often the 
manual inspection of phylogenies and sequence alignments
by volunteers~\cite{Smith2023tracking}.

@@

\subsection*{Quality control of recombination events}
Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions \cite{DeMaio2020issues} 
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
We developed a quality control (QC) measure 
which counts the number of ``loci'' (clusters of closely spaced sites)
supporting the left and right parents of a recombinant (STAR Methods),
and effectively identifies a range of issues. 
Figure~\ref{fig:recombinant_qc} plots the net number of supporting
loci for the left and right parents for all 855 recombination
events. The 501 recombinants with fewer than 4 net loci supporting 
each parent are highly enriched for the AmpliSeq V1 sequencing
protocol [TODO concrete numbers?], 
and 71 (14\%) are associated with 6-base deletion in the 
Delta lineage, known to be affected 
by amplicon dropout\cite{Sanderson2021}.
The 354 remaining recombination events that pass this QC filter
contain all of the recombinants associated with Pango X lineages
(Table~\ref{tab:pango_x_lineages}) and 
the Jackson et al recombinants (Document\ref{sec:jackson_recombs}).

@@

The LS HMM at the heart of sc2ts provides a simple and powerful means 
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution for the 
putative recombinant sequence with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between 
the two solutions---the number of 
mutations ``averted'' by recombination---is a measure of the evidence
in favour of recombination. 
Table~\ref{tab:pango_x_lineages} shows
the number of averted mutations for each of the 21 recombination events
associated with Pango X lineages, with support ranging from very
strong in the case of XC (the most parsimonious non-recombinant 
placement would require 37 additional mutations) to marginal 
for XG and XW, which have the minimal possible number of averted 
mutations.

@@

Figure~\ref{fig:averted_mutations}A plots the number of mutations required
in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
and the resulting distribution of the number of 
averted mutations (lower histogram)
over all 354 recombination events that pass QC. 
Of these, only 24 
(21 associated with Pango X lineages in Table~\ref{tab:pango_x_lineages},
and 3 additional from Jackon et al in Table~\ref{tab:jackson}) have 
% Slightly risky, this, as we could get asked to incorporate 
% novel recombs found by RIPPLEs? This could be done if we could get 
% Viridian sample IDs for their list.
been thoroughly characterised. 
The figure also shows the classification status for each recombinant
according to rebar, a method that classifies sequences
as recombinant or not based on groups of mutations associated with
Pango lineages (STAR Methods). 
Of the 38 recombinants that are strongly supported (at least 10 averted
mutations), 29 are novel. The known recombinants are XA, XC, XS, XBR, 
and 5 from Jackson et al. All but 2 of the 38 are classified as
recombinant by rebar. The majority (20) of these novel strongly supported
recombinants are singletons, with 9 having further descendants in the ARG.
Of the remaining 316 recombinants with less than 10 averted mutations,
only 151 are classified as recombinant by rebar. 
While some are likely to 
be false positives, it is also interesting to note that rebar validation
rate is much lower for recombinants with closely related 
parents (Figure~\ref{fig:averted_mutations}B,C) suggesting that 
sc2ts may have a higher sensitivity.

Recombination is rare in SARS-CoV-2, and ultimately judgements about
plausibility require expert analysis. 
To aid such interpretation and help identify quality control
issues we developed a visualisation 
based on SnpIt~\cite{OToole2024} and 
the recombinant view from RIVET~\cite{Smith2023tracking},
showing the informative genomic positions between 
a recombinant and its parent sequences, and highlights 
potentially problematic loci.
Each copying pattern shows the positions
where the allelic state in a recombinant
(middle row) matches that of the left (upper row, coloured green) 
or right parent (lower row, coloured blue),
or where the recombination event requires a de-novo mutation
(gold, with mutational change below).
Figure~\ref{fig:averted_mutations}D shows some example copying patterns,
illustrating this spectrum of plausibility.
XA is an example of a highly plausible recombinant, shown by the 
large number of supporting loci on the left and right. As we have 
only one de-novo mutation, we can be sure that the parents are closely
observed [Some details about XA here, how well the parents were sampled etc].
XBB is also strongly supported by many loci on the left and right, 
but the presence of many de-novo mutations
observed in the copying patterns suggests the existence of
many genetically distinct, unsampled ancestors 
[FIXME: in this case, probably BJ.1, see issue 336],
and some caution should be exercised in interpreting them. 
See Figures~\ref{fig:averted_mutations}D and
\ref{fig:pango_x_copying_patterns} for copying patterns of the 
Pango X lineages, and 
Figure~\ref{fig:copying_pattern_examples} for copying patterns
highlighting specific types of QC problem (and more details on the 
visualisation).

Two further recombinants in the sc2ts ARG that are likely to be 
spurious are important to discuss.
[FIXME - discuss BA.5 (kept as a recombination event but only because we can't 
rewire it post-hoc, owning to subsequent tree building) and the BA.2 event (doesn't
pass QC.]

@@

We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene (Figure~\ref{fig:recombinant_intervals}A,B),
consistent with patterns observed in previous studies
\cite{Bobay2020recombination,Turakhia2022,Smith2023tracking,Li2024CovRecomb}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
allowing recombination endpoints to be estimated more precisely.
